---
title: "Type 1 error and bias simulations for Gsens"
author: "Leo Frach"
date: "2024-08-13"
output: pdf_document
---

Simulations for power analysis comparing Gsens with and without the new correlation = TRUE feature in lavaan, as well as with and without the loading = TRUE option in Gsens (using reliability instead of heritability constraint).

## Load libraries

```{r message=FALSE}
#remotes::install_github("LeonardFrach/Gsens")
#library(Gsens)
library(lavaan)
library(stringr)
library(tidyr)
library(simstandard)
```


## Define parameters of interest

```{r}

n <- c(3000, 10000)
h2 <- c(0.20, 0.40)
b1 <- c(0, 0.30)
b2 <- c(0, 0.15)
a1 <- c(0, 0.10)
a2 <- c(0, 0.20)
pme <- c(0.50, 0.80)

cor12 <- c(0, 0.20)

# create all possible combinations
combinations <- crossing(n, b1, b2, a1, a2, h2, cor12, pme)


# add path coefficient c (effect from GF on Y), which depends on the other indirect effects of GF to Y, on h2 and on measurement error
combinations$c <-  round(with(combinations, sqrt(h2) - (a1*b1 + a2*b2)), 6) # effect of GF on outcome, half the effect of the total heritability 

combinations <- combinations %>% dplyr::filter(b1 == 0 | a1 == 0 | a2 == 0 | b2 == 0)

```

## Simulate Gsens model

```{r}
# empty lists for outputs
estimates_simu <- vector(mode = "list", length = dim(combinations)[1])
alpha_simu <- vector(mode = "list", length = dim(combinations)[1])

# create list of population models
population.modelGF <- vector(mode = "list", length = dim(combinations)[1])

myData <- vector(mode = "list", length = length(population.modelGF))
myDataX <- vector(mode = "list", length = length(population.modelGF))
```

## Population model

```{r}
# Simulate the true population models for different combinations of h2, b1, a1 and n
for (j in 1:length(population.modelGF)) {
    set.seed(123)
    population.modelGF[[j]] <-
        str_glue('                     
              # paths and loading
              Y ~ {combinations$b1[j]}*X1 + {combinations$b2[j]}*X2 + {combinations$c[j]}*GF
              X1 ~ {combinations$a1[j]}*GF
              X2 ~ {combinations$a2[j]}*GF 

              X1 ~~ {combinations$cor12[j]}*X2
              
              GF =~ sqrt({combinations$lg2[j]})*G
           ')
}


```

# Type 1 error simulations

## Simulate standard Gsens model (heritability) *without* correlation = TRUE feature

```{r}
Sys.time()

output <- NULL

set.seed(12345)

for (j in 1:length(population.modelGF)) {
     k = 1000
     # create empty lists
     # myData <- vector(mode = "list", length = k)
     # myDataX <- vector(mode = "list", length = k)
     # output <- vector(mode = "list", length = k)
    
     for (i in 1:k) {
         myData[[i]] <- sim_standardized(population.modelGF[[j]], n = combinations$n[j],
                                         latent = TRUE,
                                         errors = TRUE)
    
         # Specify Gsens model (note latent variable here is GG to avoid confusion with simulated GF that is in the dataset).
         myDataX[[i]] <- myData[[i]][, 1:4]
    
         # run gsensY function from script gsens.R
         output[[i]] <- gsensY(sample.cov = cov(myData[[i]]),
                               sample.nobs = combinations$n[j],
                               h2 = combinations$h2[j],
                               exposures = c("X1", "X2"),
                               outcome = "Y", pgs = "G",
                               estimator = "GLS",
                               loading = FALSE)
     }


    # get average estimates and standard errors for all simulated data sets
    
    sum_est = 0
    sum_se = 0
    
    z = rep(k, 9)
    
    for (i in 1:k) {
        
        z_est <- z - is.na(output[[i]]@external$gsensY$est)
        z_se <- z - is.na(output[[i]]@external$gsensY$se)

        sum_est <- output[[i]]@external$gsensY$est + sum_est
        sum_se <- output[[i]]@external$gsensY$se + sum_se
        
    }
    
    mean_est <- sum_est / z_est
    mean_se <- sum_se / z_se
    
    
    estimates <- cbind(mean_est, mean_se)
    rownames(estimates) <- rownames(output[[i]]@external$gsensY)
    
    # power
    count_p <- NULL
    
    for (i in 1:k) {
        count_p <- cbind(output[[i]]@external$gsensY$pvalue, count_p)
    }
    
    
    alpha <- NULL
    
    for (i in 1:dim(count_p)[1]) {
        alpha[i] <- mean(as.numeric(count_p[i, ]) < 0.05, na.rm = T)
    }
    
    estimates_simu[[j]] <- estimates
    alpha_simu[[j]] <- round(alpha, 3)
}

Sys.time()

alpha_simu
estimates_simu

# add to combinations df
for (i in 1:length(alpha_simu)) {
    combinations$alpha_b1[i] <- alpha_simu[[i]][1] %>% as.numeric()
    combinations$alpha_b2[i] <- alpha_simu[[i]][2] %>% as.numeric()
    combinations$alpha_m1[i] <- alpha_simu[[i]][3] %>% as.numeric()
    combinations$alpha_m2[i] <- alpha_simu[[i]][4] %>% as.numeric()
    combinations$alpha_m_total[i] <- alpha_simu[[i]][5] %>% as.numeric()
    combinations$alpha_gc1[i] <- alpha_simu[[i]][6] %>% as.numeric()
    combinations$alpha_gc2[i] <- alpha_simu[[i]][7] %>% as.numeric()
}

for (i in 1:length(estimates_simu)) {
    combinations$est_b1[i] <- round(estimates_simu[[i]][1] %>% as.numeric(), 3)
    combinations$est_b2[i] <- round(estimates_simu[[i]][2] %>% as.numeric(), 3)
    combinations$est_m1[i] <- round(estimates_simu[[i]][3] %>% as.numeric(), 3)
    combinations$est_m2[i] <- round(estimates_simu[[i]][4] %>% as.numeric(), 3)
    combinations$est_m_total[i] <- round(estimates_simu[[i]][5] %>% as.numeric(), 3)
    combinations$est_gc1[i] <- round(estimates_simu[[i]][6] %>% as.numeric(), 3)
    combinations$est_gc2[i] <- round(estimates_simu[[i]][7] %>% as.numeric(), 3)
}


combinations$bias_b1 <- combinations$est_b1 - combinations$b1 
combinations$bias_b2 <- combinations$est_b2 - combinations$b2
combinations$bias_m1 <- combinations$est_m1 - (combinations$a1 * combinations$b1)
combinations$bias_m2 <- combinations$est_m2 - (combinations$a2 * combinations$b2) 
combinations$bias_m_total <- combinations$est_m_total - ((combinations$a1 * combinations$b1) + (combinations$a2 * combinations$b2)) 

combinations$bias_gc1 <- combinations$est_gc1 - ((combinations$a1 * combinations$c) + (combinations$a2 * combinations$b2)) 

combinations$bias_gc2 <- combinations$est_gc2 - ((combinations$a2 * combinations$c) + (combinations$a1 * combinations$b1))

save(combinations, file = "alpha.RData")


```

# Plot type 1 error and bias
```{r}
library(ggplot2)

# add info on effect sizes
df <- na.omit(combinations) 
#df <- pivot_longer()

df_alpha_b1 <- df %>% 
    dplyr::filter(b1 == 0) %>%
    dplyr::mutate(pme = as.factor(pme),
                  cor12 = as.factor(cor12),
                  h2 = as.factor(h2))

df_a_b1 <- df_alpha_b1 %>% group_by(pme, n) %>%
    summarise(alpha = mean(alpha_b1))

library(wesanderson)
library(colorBlindness)


colorBlindness::availablePalette()
colorBlindness::displayAllColors(paletteMartin)

cols <- paletteMartin
cols
# Black    SherpaBlue  PersianGreen       HotPink   CottonCandy PigmentIndigo   ScienceBlue    Heliotrope        Malibu    FrenchPass      RedBerry 
#cols <- c("#000000",  "#009292" , "#ff6db6"  ,  "#490092", "#db6d00" ) 
cols <- c( "#490092", "#db6d00" ) 

df_alpha_b2 <- df %>% 
    dplyr::filter(b2 == 0) %>%
    dplyr::mutate(pme = as.factor(pme),
                  cor12 = as.factor(cor12),
                  h2 = as.factor(h2))

df_a_b2 <- df_alpha_b2 %>% group_by(pme, n) %>%
    summarise(alpha = mean(alpha_b2))


df_alpha_m1 <- df %>% 
    dplyr::filter(b1 == 0 | a1 == 0) %>%
    dplyr::mutate(pme = as.factor(pme),
                  cor12 = as.factor(cor12),
                  h2 = as.factor(h2))

df_a_m1 <- df_alpha_m1 %>% group_by(pme, n) %>%
    summarise(alpha = mean(alpha_m1))

df_alpha_m2 <- df %>% 
    dplyr::filter(a2 == 0 | b2 == 0) %>%
    dplyr::mutate(pme = as.factor(pme),
                  cor12 = as.factor(cor12),
                  h2 = as.factor(h2))

df_a_m2 <- df_alpha_m2 %>% group_by(pme, n) %>%
    summarise(alpha = mean(alpha_m2))


df_alpha_gc1 <- df %>% 
    dplyr::filter(a1 == 0) %>%
    dplyr::mutate(pme = as.factor(pme),
                  cor12 = as.factor(cor12),
                  h2 = as.factor(h2))

df_a_gc1 <- df_alpha_gc1 %>% group_by(pme, n) %>%
    summarise(alpha = mean(alpha_gc1))


df_alpha_gc2 <- df %>% 
    dplyr::filter(a2 == 0) %>%
    dplyr::mutate(pme = as.factor(pme),
                  cor12 = as.factor(cor12),
                  h2 = as.factor(h2))

df_a_gc2 <- df_alpha_gc2 %>% group_by(pme, n) %>%
    summarise(alpha = mean(alpha_gc2))


png(filename = "alpha_error_h2_Gsens.png", width = 4000, height = 4000, type = "cairo", res = 600)
alpha_b1 <- ggplot(df_a_b1, 
                    aes(x = as.factor(n),
                        y = alpha,
                        fill = pme)) + 
    geom_bar(aes(fill = pme), position = "dodge", stat = "identity") + 
    geom_hline(yintercept = 0.05, linetype = 2) + 
    scale_y_continuous(breaks = seq(0, 0.15, 0.01), limits = c(0, 0.15)) + #this line change the frequency of tick marks
    scale_fill_manual(values =  cols) +
    labs(fill = "Measurement error", x = "Sample size", y = "Type 1 error") 

alpha_b2 <- ggplot(df_a_b2, 
                    aes(x = as.factor(n),
                        y = alpha,
                        fill = pme)) + 
    geom_bar(aes(fill = pme), position = "dodge", stat = "identity") + 
    geom_hline(yintercept = 0.05, linetype = 2) + 
    scale_y_continuous(breaks = seq(0, 0.15, 0.01), limits = c(0, 0.15)) + #this line change the frequency of tick marks
    scale_fill_manual(values =  cols) +
    labs(fill = "Measurement error", x = "Sample size", y = "Type 1 error") 

alpha_m1 <- ggplot(df_a_m1, 
                    aes(x = as.factor(n),
                        y = alpha,
                        fill = pme)) + 
    geom_bar(aes(fill = pme), position = "dodge", stat = "identity") + 
    geom_hline(yintercept = 0.05, linetype = 2) + 
    scale_y_continuous(breaks = seq(0, 0.15, 0.01), limits = c(0, 0.15)) + #this line change the frequency of tick marks
    scale_fill_manual(values =  cols) +
    labs(fill = "Measurement error", x = "Sample size", y = "Type 1 error") #+
    #theme(legend.position = "none")

alpha_m2 <- ggplot(df_a_m2, 
                    aes(x = as.factor(n),
                        y = alpha,
                        fill = pme)) + 
    geom_bar(aes(fill = pme), position = "dodge", stat = "identity") + 
    geom_hline(yintercept = 0.05, linetype = 2) + 
    scale_y_continuous(breaks = seq(0, 0.15, 0.01), limits = c(0, 0.15)) + #this line change the frequency of tick marks
    scale_fill_manual(values =  cols) +
    labs(fill = "Measurement error", x = "Sample size", y = "Type 1 error")

alpha_gc1 <- ggplot(df_a_gc1, 
                    aes(x = as.factor(n),
                        y = alpha,
                        fill = pme)) + 
    geom_bar(aes(fill = pme), position = "dodge", stat = "identity") + 
    geom_hline(yintercept = 0.05, linetype = 2) + 
    scale_y_continuous(breaks = seq(0, 0.15, 0.01), limits = c(0, 0.15)) + #this line change the frequency of tick marks
    scale_fill_manual(values =  cols) +
    labs(fill = "Measurement error", x = "Sample size", y = "Type 1 error") 

alpha_gc2 <- ggplot(df_a_gc2, 
                    aes(x = as.factor(n),
                        y = alpha,
                        fill = pme)) + 
    geom_bar(aes(fill = pme), position = "dodge", stat = "identity") + 
    geom_hline(yintercept = 0.05, linetype = 2) + 
    scale_y_continuous(breaks = seq(0, 0.15, 0.01), limits = c(0, 0.15)) + #this line change the frequency of tick marks
    scale_fill_manual(values =  cols) +
    labs(fill = "Measurement error", x = "Sample size", y = "Type 1 error") 

cowplot::plot_grid(
    cowplot::plot_grid(alpha_b1 + theme(legend.position = "none"), 
                       alpha_b2 + theme(legend.position = "none"),
                       alpha_m1 + theme(legend.position = "none"), 
                       alpha_m2 + theme(legend.position = "none"), 
                       alpha_gc1 + theme(legend.position = "none"), 
                       alpha_gc2 + theme(legend.position = "none"),
                       ncol = 2,
                       labels = c("a", "b", "c", "d", "e", "f")),
    cowplot::get_legend(alpha_b1 + scale_shape(guide = FALSE) + 
                            theme(legend.position = "right")),
    nrow = 1, rel_widths = c(.80, .20))


dev.off()
```



